---
title: "Exploratory Data Analysis - Human Microbiome Project (5 Body Sites)"
author: "Kexin Cui"
date: "2026-04-19"
output: html_document
---

# Overview

This Quarto file performs exploratory data analysis (EDA) on a subset of the
Human Microbiome Project (HMP) 16S rRNA V3–V5 dataset. The goal of the EDA is
to describe the structure of the processed microbiome data and to give an
initial look at whether communities differ across the five major body sites
(Airways, Gut, Oral, Skin, Urogenital). The EDA here motivates the formal
statistical analyses that are performed in
`code/analysis-code/statistical-analysis.R`.

The three hypotheses guiding the project are:

- H1: Microbial alpha diversity differs significantly across the five major
  body sites.
- H2: Microbial communities cluster distinctly by body site in beta diversity
  ordination space.
- H3: Microbial co-occurrence networks differ in structure across body sites
  (topology, modularity, hub taxa).

# Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(HMP16SData)
library(phyloseq)
library(vegan)
library(here)
```

```{r}
# Consistent colors for the five major body sites
colors_body <- c(
  "Airways"    = "#87CEEB",
  "Gut"        = "#8B4513",
  "Oral"       = "#DC143C",
  "Skin"       = "#FFB6C1",
  "Urogenital" = "#9370DB"
)
theme_set(theme_bw(base_size = 12))
```

# Load data

We use the processed phyloseq objects saved by the processing step. `ps_filt`
contains filtered counts and `ps_rel` contains relative abundances.

```{r}
ps_filt <- readRDS(here("data", "processed-data", "ps_filt.rds"))
ps_rel  <- readRDS(here("data", "processed-data", "ps_rel.rds"))

# Use the body-site simplification from the processing step. If SampleType
# is not already available, create it from HMP_BODY_SITE.
if (!"SampleType" %in% colnames(sample_data(ps_filt))) {
  sample_data(ps_filt)$SampleType <- sample_data(ps_filt)$HMP_BODY_SITE |>
    gsub("Gastrointestinal Tract", "Gut", x = _) |>
    gsub("Urogenital Tract", "Urogenital", x = _)
  sample_data(ps_rel)$SampleType <- sample_data(ps_filt)$SampleType
}
```

# Sequencing depth

Sequencing depth varies substantially across samples, which motivates the
rarefaction step in the processing pipeline.

```{r}
p_seq <- tibble(library_size = sample_sums(ps_filt)) |>
  ggplot(aes(x = library_size)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(x = "Sequencing depth (reads per sample)",
       y = "Number of samples")

ggsave(here("results", "figures", "seq_depth.png"),
       p_seq, width = 6, height = 4, dpi = 300)
p_seq
```

# Dominant phyla

Stacked bars of the 10 most abundant phyla show that community composition is
dominated by a small set of phyla that vary across samples — consistent with
previously described HMP patterns.

```{r}
otu <- as(otu_table(ps_rel), "matrix")
if (taxa_are_rows(ps_rel)) otu <- t(otu)

tax <- as(tax_table(ps_rel), "matrix")
phylum_col <- if ("Phylum" %in% colnames(tax)) "Phylum" else "PHYLUM"
phylum <- tax[, phylum_col]
phylum[is.na(phylum) | phylum == ""] <- "UNASSIGNED"

otu_phylum <- t(rowsum(t(otu), group = phylum))

df <- as_tibble(otu_phylum, rownames = "SampleID") |>
  pivot_longer(-SampleID, names_to = "Phylum", values_to = "RelAbund")

top10 <- df |>
  group_by(Phylum) |>
  summarize(total = sum(RelAbund), .groups = "drop") |>
  slice_max(total, n = 10) |>
  pull(Phylum)

p_phylum <- df |>
  filter(Phylum %in% top10) |>
  ggplot(aes(SampleID, RelAbund, fill = Phylum)) +
  geom_col() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "Samples", y = "Relative abundance")

ggsave(here("results", "figures", "top_phyla.png"),
       p_phylum, width = 8, height = 4, dpi = 300)
p_phylum
```

# Alpha diversity across the five body sites (H1)

We compute four alpha diversity indices (Observed richness, Shannon, Simpson,
Chao1) on the filtered counts and test for differences across body sites using
non-parametric Kruskal-Wallis tests. Boxplots summarize within-site diversity.

```{r}
# Alpha diversity should be calculated on integer counts
alpha_div <- estimate_richness(
  ps_filt,
  measures = c("Observed", "Shannon", "Simpson", "Chao1")
)
alpha_div$SampleType <- sample_data(ps_filt)$SampleType

metrics <- c("Observed", "Shannon", "Simpson", "Chao1")
kw_results <- map_dfr(metrics, function(m) {
  kw <- kruskal.test(as.formula(paste(m, "~ SampleType")), data = alpha_div)
  tibble(Metric = m,
         ChiSq  = unname(kw$statistic),
         df     = unname(kw$parameter),
         p      = kw$p.value)
})

kw_results
```

```{r}
alpha_long <- alpha_div |>
  pivot_longer(all_of(metrics), names_to = "Metric", values_to = "Value")

p_alpha <- ggplot(alpha_long,
                  aes(x = SampleType, y = Value, fill = SampleType)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  facet_wrap(~Metric, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = colors_body) +
  labs(title = "Alpha diversity across five major body sites",
       x = "Body Site", y = "Diversity value") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("results", "figures", "alpha_diversity_5sites.png"),
       p_alpha, width = 10, height = 8, dpi = 300)
p_alpha
```

The Shannon boxplot alone (kept from the earlier version for comparison):

```{r}
p_shannon <- alpha_div |>
  ggplot(aes(x = SampleType, y = Shannon, fill = SampleType)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = colors_body) +
  labs(x = "Body Site", y = "Shannon diversity") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("results", "figures", "shannon_boxplot.png"),
       p_shannon, width = 7, height = 4, dpi = 300)
p_shannon
```

# Beta diversity / ordination (H2, preview)

As a preview of the formal PERMANOVA analysis in `statistical-analysis.R`, we
run a PCoA on log(1+x)-transformed counts with Bray-Curtis distance.

```{r}
ps_log <- transform_sample_counts(ps_filt, function(x) log(1 + x))
ord_bc <- ordinate(ps_log, method = "PCoA", distance = "bray")
evals <- ord_bc$values$Eigenvalues
var1  <- round(100 * evals[1] / sum(evals), 1)
var2  <- round(100 * evals[2] / sum(evals), 1)

ord_df <- data.frame(ord_bc$vectors[, 1:2])
ord_df$SampleType <- sample_data(ps_log)$SampleType

p_pcoa <- ggplot(ord_df, aes(Axis.1, Axis.2, color = SampleType)) +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  scale_color_manual(values = colors_body) +
  labs(x = sprintf("PC1 (%.1f%%)", var1),
       y = sprintf("PC2 (%.1f%%)", var2),
       color = "Body Site",
       title = "PCoA (Bray-Curtis) - preview")

ggsave(here("results", "figures", "pcoa_preview.png"),
       p_pcoa, width = 8, height = 6, dpi = 300)
p_pcoa
```

# Summary table by body site

```{r}
explore_tab <- alpha_div |>
  rownames_to_column("SampleID") |>
  mutate(library_size = sample_sums(ps_filt)[SampleID]) |>
  group_by(SampleType) |>
  summarize(
    n_samples         = n(),
    mean_library_size = mean(library_size, na.rm = TRUE),
    mean_observed     = mean(Observed,     na.rm = TRUE),
    mean_shannon      = mean(Shannon,      na.rm = TRUE),
    sd_shannon        = sd(Shannon,        na.rm = TRUE),
    mean_simpson      = mean(Simpson,      na.rm = TRUE),
    .groups = "drop"
  )

saveRDS(explore_tab,
        here("results", "tables", "exploratory_summary_by_subsite.rds"))
explore_tab
```

# Take-aways from EDA

- Sequencing depth is variable; rarefaction in the processing step normalizes
  depth before downstream analyses.
- A small number of phyla dominate the community, with visible between-sample
  differences that track body site.
- Alpha diversity boxplots show clear visual separation across the five major
  body sites, and Kruskal-Wallis tests above will quantify this in H1.
- The PCoA preview shows samples tending to cluster by body site, motivating
  the formal PERMANOVA in `statistical-analysis.R` (H2).
- Network construction (H3) is performed per body site in the analysis script.
