---
title: "An example exploratory analysis script"
date: "2024-02-07"
output: html_document
---

  
  
This Quarto file loads the cleaned data and does some exploring.

This version has the code inside the Quarto file itself.

Also note that while cleaning and exploring is split here, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. 

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# BiocManager::install("microbiome")

library(tidyverse)
library(HMP16SData)
library(phyloseq)
library(vegan)
library(microbiome)
library(here)
```


Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ps_rel.rds")
#load data
mydata <- readRDS(data_location)
```







# Data exploration through figures

```{r}
ps_filt <- readRDS(here("data","processed-data","ps_filt.rds"))

p_seq <- tibble(library_size = sample_sums(ps_filt)) %>%
  ggplot(aes(x = library_size)) +
  geom_histogram(bins = 30) +
  labs(x = "Sequencing depth (reads per sample)",
       y = "Number of samples")

ggsave(here("results","figures","seq_depth.png"),
       p_seq, width = 6, height = 4, dpi = 300)
```

Dominant taxa composition (top phyla)
```{r}
# Extract relative abundance matrix
otu <- as(otu_table(mydata), "matrix")
if (taxa_are_rows(mydata)) otu <- t(otu)

tax <- as(tax_table(mydata), "matrix")
phylum <- tax[, "PHYLUM"]
phylum[is.na(phylum) | phylum == ""] <- "UNASSIGNED"

otu_phylum <- rowsum(t(otu), group = phylum)
otu_phylum <- t(otu_phylum)

df <- as_tibble(otu_phylum, rownames = "SampleID") |>
  pivot_longer(-SampleID, names_to = "PHYLUM", values_to = "RelAbund")

top10 <- df |>
  group_by(PHYLUM) |>
  summarize(total = sum(RelAbund), .groups = "drop") |>
  slice_max(total, n = 10) |>
  pull(PHYLUM)

p_phylum <- df |>
  filter(PHYLUM %in% top10) |>
  ggplot(aes(SampleID, RelAbund, fill = PHYLUM)) +
  geom_col() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "Samples", y = "Relative abundance")

ggsave(
  filename = here("results","figures","top_phyla.png"),
  plot = p_phylum,
  width = 8,
  height = 4,
  dpi = 300
)
```

Relative abundances of dominant bacterial phyla varied substantially across samples, indicating heterogeneous community composition typical of human microbiome data.

Alpha diversity (Shannon) + boxplot by group
```{r}
alpha <- estimate_richness(mydata, measures = "Shannon") %>%
  rownames_to_column("SampleID")

meta <- as(sample_data(mydata), "data.frame") %>%
  rownames_to_column("SampleID")

df_alpha <- left_join(alpha, meta, by = "SampleID")

GROUP_VAR <- "HMP_BODY_SUBSITE"

p_alpha <- df_alpha %>%
  ggplot(aes(x = .data[[GROUP_VAR]], y = Shannon)) +
  geom_boxplot() +
  labs(x = GROUP_VAR, y = "Shannon diversity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = here("results","figures","shannon_boxplot.png"),
  plot = p_alpha,
  width = 7,
  height = 4,
  dpi = 300
)
```

# Data exploration through tables

```{r}
GROUP_VAR <- "HMP_BODY_SUBSITE"

explore_tab <- df_alpha %>%
  mutate(
    library_size = sample_sums(mydata)[SampleID],
    group = .data[[GROUP_VAR]]
  ) %>%
  group_by(group) %>%
  summarize(
    n_samples = n(),
    mean_library_size = mean(library_size, na.rm = TRUE),
    mean_shannon = mean(Shannon, na.rm = TRUE),
    sd_shannon = sd(Shannon, na.rm = TRUE),
    .groups = "drop"
  )

# save only as RDS
saveRDS(
  explore_tab,
  file = here("results","tables","exploratory_summary_by_subsite.rds")
)
```



# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.


