---
title: "An example cleaning script"
author: "Andreas Handel"
date: "2024-02-07"
output: html_document
---


# Processing script

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.


# Setup

Load needed packages. make sure they are installed.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# BiocManager::install("microbiome")

library(tidyverse)
library(HMP16SData)
library(phyloseq)
library(vegan)
library(microbiome)
library(here)
```


# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}


# Load Human Microbiome Project 16S data (V3–V5 region)
se <- V35()
ps <- as_phyloseq(se)


```


# Check data

First we can look at the codebook

```{r}
# Basic dimensions
nsamples(ps)
ntaxa(ps)

# Available sample metadata
sample_variables(ps)
```



# Cleaning

By inspecting the data as done above, we find some problems that need addressing:

```{r}
# Inspect taxonomy and metadata
tax_table(ps)[1:5, 1:4]
sample_data(ps)[1:5, ]

# Remove taxa with zero total counts
ps_nz <- prune_taxa(taxa_sums(ps) > 0, ps)

# Filter rare taxa (present in >5 samples)
ps_filt <- filter_taxa(ps_nz, function(x) sum(x > 0) > 5, prune = TRUE)

# Transform counts to relative abundance
ps_rel <- transform_sample_counts(ps_filt, function(x) x / sum(x))

# Compare dimensions before and after filtering
tibble(
  stage = c("Original", "Filtered"),
  samples = c(nsamples(ps), nsamples(ps_rel)),
  taxa = c(ntaxa(ps), ntaxa(ps_rel))
)
```


# Save data 

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
saveRDS(ps_filt, here::here("data","processed-data","ps_filt.rds"))
saveRDS(ps_rel,  here::here("data","processed-data","ps_rel.rds"))
```



# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

