---
title: "Supplementary Material - Comparing Microbial Diversity and Co-Occurrence Networks Across Five Major Body Sites (HMP 16S)"
author: "Kexin Cui"
date: today
format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../../assets/dataanalysis-references.bib
csl: ../../../assets/vancouver-author-date.csl
---

```{r, echo=FALSE, message=FALSE}
library(here)
library(knitr)
library(dplyr)
```

# Overview

This supplement documents (a) the full set of files and the order in which
they must be run to reproduce every figure and table in the main manuscript,
(b) additional methodological detail, and (c) extra results that did not fit
into the main text (pairwise alpha diversity comparisons, detailed network
summaries).

# Code and file information

The project has three code stages. Run them in the following order:

1. `code/processing-code/processingcode.R` *(or `processingfile-v2.qmd`)* —
   downloads the HMP V3–V5 data with `HMP16SData::V35()`, performs stratified
   subsampling, prevalence and abundance filtering, and writes
   `data/processed-data/ps_filt.rds` and `data/processed-data/ps_rel.rds`.
2. `code/eda-code/eda.qmd` — produces the exploratory figures
   (`seq_depth.png`, `top_phyla.png`, `alpha_diversity_5sites.png`,
   `shannon_boxplot.png`, `pcoa_preview.png`) and the summary table
   `results/tables/exploratory_summary_by_subsite.rds`.
3. `code/analysis-code/statistical-analysis.R` — runs the formal hypothesis
   tests (Kruskal-Wallis, pairwise Wilcoxon, PERMANOVA, BETADISPER),
   builds per-site co-occurrence networks, and writes every table used in
   the manuscript (`alpha_kruskal.rds`, `alpha_pairwise.rds`,
   `permanova.rds`, `network_stats.rds`) plus figures
   (`pcoa_ordination_5sites.png`, `hierarchical_clustering_5sites.png`,
   `network_statistics_comparison.png`).

The manuscript Quarto file `products/manuscript/Manuscript.qmd` pulls these
results in via `readRDS()` and `knitr::include_graphics()`.

{{< pagebreak >}}

# Additional method details

## Stratified subsampling

To make the comparison across the five major body sites balanced, we sample
up to 200 samples per site using base R's `sample()` with `set.seed(123)`.
Sites with fewer than 200 samples are included in full. This procedure gives
the same set of samples on every run.

## Filtering and rarefaction

Taxa are retained only if they are present in at least 10 samples and have a
total count of ≥ 50 reads. Samples are rarefied to the 5th percentile of the
distribution of sample sums, using `phyloseq::rarefy_even_depth()` with
`rngseed = 123`. This threshold keeps 95% of samples while substantially
reducing variation in sequencing depth.

## Distance choice

We use Bray-Curtis as the primary distance because it is abundance-aware
and is the default distance in most HMP-era papers. Jaccard (presence/
absence) is used as a robustness check. BETADISPER tests whether the
PERMANOVA effect is confounded with unequal dispersion among groups; both
tests are reported.

## Network construction

Correlation networks are built per body site from the prevalence-filtered
OTU table (≥ 10% prevalence) using Spearman correlations via `Hmisc::rcorr`.
Edges are retained when |r| ≥ 0.5 and p < 0.01 (no multiple-testing
correction is applied inside the network — this is a pragmatic threshold,
not a formal significance test). Louvain community detection
(`igraph::cluster_louvain`) is used to compute modularity and communities.

{{< pagebreak >}}

# Additional results

## Pairwise alpha diversity comparisons

Table @tbl-alpha-pairwise lists the Benjamini-Hochberg-adjusted
pairwise Wilcoxon p-values for every pair of body sites, for each of the
four alpha-diversity indices. These follow up the overall Kruskal-Wallis
tests reported in the main text.

```{r, echo=FALSE}
#| label: tbl-alpha-pairwise
#| tbl-cap: "Pairwise Wilcoxon (BH-adjusted) p-values for alpha diversity differences across body sites."
pw_path <- here::here("results", "tables", "alpha_pairwise.rds")
if (file.exists(pw_path)) {
  pw <- readRDS(pw_path)
  knitr::kable(pw, digits = 4)
} else {
  cat("Run statistical-analysis.R to produce this table.")
}
```

## Full network statistics

Table @tbl-network-full provides the full per-site network statistics
(reproduced here from the main text for convenience, together with the
number of positive vs. negative edges that are not shown in the main
figure).

```{r, echo=FALSE}
#| label: tbl-network-full
#| tbl-cap: "Per-site co-occurrence network summary."
net_path <- here::here("results", "tables", "network_stats.rds")
if (file.exists(net_path)) {
  ns <- readRDS(net_path)
  knitr::kable(ns, digits = 3)
} else {
  cat("Run statistical-analysis.R to produce this table.")
}
```

## Individual network visualizations

The networks themselves are saved as an `igraph` list in
`results/output/networks.rds`. Each network can be loaded individually and
visualized with `ggraph`; below is a placeholder for a figure panel showing
one representative network per body site.

```{r, echo=FALSE}
#| label: fig-network-oral
#| fig-cap: "Representative co-occurrence network visualization (replace with a saved network PNG)."
fig_path <- here::here("results", "figures",
                       "network_statistics_comparison.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

{{< pagebreak >}}

# Discussion

Results in this supplement support the three main-text conclusions: (1) the
overall Kruskal-Wallis tests are not driven by a single outlier site — most
pairwise contrasts are significant after BH correction; (2) PERMANOVA
effects are robust to switching between Bray-Curtis and Jaccard distances;
and (3) network differences across sites are not artifacts of sample size
alone, because the ranking of sites by modularity and communities is
consistent with the ecological structure reported in the literature
[@mckay2020; @mckay2020a].

{{< pagebreak >}}

# References
